---
title: |
  <div style="text-align:center;">Project Report: COVID-19 Analysis in Mexico</div>
subtitle: |
  <div style="text-align:center;">DATA 602: Statistical Data Analysis</div>
author: |
  <div style="text-align:center;">Kanwarpal Singh (30232373), Faiq Mohd Zulkarnain (30207865), Osama Bilgrami (30213085), Arteen Rafiei (30043409)</div>
output: 
  html_document:
    theme: cosmo
    toc: true
    toc-title: "Table of Contents"
editor_options: 
  markdown: 
    wrap: 72
---

```{r, echo=FALSE, error=FALSE, message=FALSE}
options(scipen=999)

library(ggplot2)
library(knitr)
library(dplyr)
library(mosaic)
```

```{r, echo=FALSE, error=FALSE, message=FALSE}
set.seed(1997)
```

<br><br>

### **Introduction**

Following the tremendous challenges posed by the COVID-19 pandemic,
humanity has reached an important turning point in history, compelled to
investigate, and comprehend the complexities of pandemics and their
far-reaching implications. As we enter an era where the possibility of
future pandemics looms large, it is critical that we strengthen our
readiness and response processes. To that purpose, we are doing two
independent but interconnected statistical analyses that have the
potential to dramatically improve our capacity to predict the
consequences of such events, similar to the COVID-19 pandemic.

The fundamental goal of our analytical efforts is to reach conclusions
that will allow us to delve deeper into the complex effects of pandemics
on various aspects of human existence. Our primary focus is on the
investigation of crucial characteristics such as age, habits, and
comorbidities, which have emerged as critical determinants in the
evolution and outcome of pandemics. We hope to untangle the deep link
between these parameters and the trajectory of pandemic-related outcomes
through careful investigation and synthesis of multiple data sources.

Our first statistical test delves into hypothesis testing. Here, we aim
to determine whether individuals aged 60 and older experience a more
significant impact and higher mortality rates in comparison to those
aged 18 and below due to COVID. In this context, our null hypothesis
asserts that there is no difference in the proportion of COVID-related
deaths between individuals above 60 and those below 18, with the
difference being equal to 0. Conversely, our alternative hypothesis
suggests that the mean difference is greater than 0, indicating a higher
impact and mortality rate among the elderly.

In our second statistical test, we focus on investigating potential
associations between an individual's number of pre-existing health
conditions and various aspects of COVID. This includes examining the
likelihood of testing positive for the virus and mortality. Moreover, we
aim to explore how comorbidities interact with factors such as age and
gender. The objective here is to unveil the relationships and
dependencies that exist between these variables.

The findings of these statistical investigations are not only
significant in and of themselves, but also have implications for
healthcare and policy. They have the potential to serve as the
foundation for developing more targeted and effective response tactics.
We hope to provide vital information to government officials, healthcare
professionals, and policymakers by identifying the persons most
vulnerable to the negative impacts of pandemics. In doing so, we hope to
give them the skills they need to make informed decisions, allocate
resources wisely, and prioritize treatments for individuals who are at a
higher risk than others.

<br>

### **Dataset**

The dataset, generously provided by the Mexican government, has been
made accessible on Kaggle and has been licensed as a CC0 Public Domain
License, allowing users unrestricted utilization. Translated by a Kaggle
user, this semi-structured dataset boasts 1,048,576 unique rows and 21
features encompassing anonymized patient-related information, including
pre-existing conditions. The data exhibit a predominantly boolean
format, utilizing 0 to denote male and 1 for female. Similarly,
pre-existing conditions are also encoded as boolean values, with 0
signifying an absence and 1 indicating a presence.

There are also presence of scaled data types ranging from 1 to 5, such
as the COVID severity classification, contributing to the dataset's
richness. Additionally, values of 97 and 99 under any feature serve as
indicators of missing data. This comprehensive dataset thus presents a
valuable resource for researchers and data enthusiasts alike, offering
insights into the intricate landscape of patient profiles and health
conditions in the Mexican context.

```{r, echo=FALSE, error=FALSE, message=FALSE, results=FALSE}
df = read.csv('coviddataset.csv')
head(df, 5)
```

<br>

### **Data Cleaning & Wrangling**

We went through several steps to clean up and prepare the data for
further analysis. Null values, which are frequently represented as
numbers, were replaced with "NA," as this prevents missing data from
distorting later analysis. Furthermore, additional features were
cleverly designed to help with deeper insights, such as the generation
of age brackets based on predetermined boundaries, segmenting the data
into '0-13,' '14-19,' '20-35,' '36-49,' and '50+' age groupings.
Standardizing data representation is crucial for ensuring consistency in
the dataset. For instance, patient types were converted from '1' and '2'
to '0' and '1', and sex was converted to 'male' and 'female'. To
facilitate more comprehensive analysis, we introduced additional
columns. One such column categorizes comorbidities and the number of
diseases. Additionally, we create a 'mortality' column that classifies
individuals as either 'Dead' or 'Alive,' and a 'covid_test_result'
column that categorizes test outcomes as 'Positive' or 'Negative.'

```{r, echo=FALSE, error=FALSE, message=FALSE}
names(df) = tolower(names(df))

#df$icu[df$icu %in% c(97, 98, 99)] = NA
df[df == 97] = NA
df[df == 98] = NA
df[df == 99] = NA

df$date_died[df$date_died == '9999-99-99'] = NA

```

```{r, echo=FALSE, error=FALSE, message=FALSE}
cols_to_replace = c('sex', 'intubed', 'pneumonia', 'pregnant', 'diabetes', 'copd', 'asthma', 'inmsupr', 'hipertension', 'other_disease', 'cardiovascular', 'obesity', 'renal_chronic', 'tobacco', 'icu')

# Replace all occurrences of 2 with 0 in the specified columns
df = df %>% mutate(across(all_of(cols_to_replace), ~ifelse(. == 2, 0, .)))
```

```{r, echo=FALSE, error=FALSE, message=FALSE}
df$sex[df$sex == 0] <- "male"
df$sex[df$sex == 1] <- "female"
```

```{r, echo=FALSE, error=FALSE, message=FALSE}
df$patient_type[df$patient_type == 1] = 0
df$patient_type[df$patient_type == 2] = 1
```

```{r, echo=FALSE, error=FALSE, message=FALSE}
breaks <- c(0, 13, 19, 35, 49, 70, 150)
labels <- c('0-13', '14-19', '20-35', '36-49', '50-70', '70+')

df$age_range <- cut(df$age, breaks = breaks, labels = labels, include.lowest = TRUE)
```

```{r, echo=FALSE, error=FALSE, message=FALSE}
cols_to_sum = c('pneumonia', 'diabetes', 'copd', 'asthma', 'inmsupr', 'hipertension', 'other_disease', 'cardiovascular', 'renal_chronic')

# Create a new column 'number_of_disease' with the sum of selected columns
df = df %>% mutate(number_of_disease = rowSums(select(., all_of(cols_to_sum))))
```

```{r, echo=FALSE, error=FALSE, message=FALSE}
df$mortality = ifelse(is.na(df$date_died), "Alive", "Dead")
```

```{r, echo=FALSE, error=FALSE, message=FALSE}
df$covid_test_result = cut(df$clasiffication_final, breaks = c(0, 3, 8), labels = c("Positive", "Negative"))
df$covid_test_result = as.factor(df$covid_test_result)
```

```{r, echo=FALSE, error=FALSE, message=FALSE}
df$covid_severity = cut(df$clasiffication_final, breaks = c(0, 1, 2, 3, 8), labels = c("High", "Medium", "Low", "Negative"))
df$covid_severity = as.factor(df$covid_severity)
```

```{r, echo=FALSE, error=FALSE, message=FALSE}
head(df, 5)
```

<br>

### **Data Analysis**

<br>

#### ***Initial Exploratory Data Analysis***

<br>

```{r, echo=FALSE, error=FALSE, message=FALSE}

ggplot(data=filter(df, (covid_test_result == "Positive")), aes(x=mortality, y=age, color = mortality, fill = mortality)) + geom_violin(na.rm=TRUE) + geom_boxplot(fill = "white", width=0.1, na.rm=TRUE) + ylab("Age") + ggtitle("Age Distribution by Mortality: Violin Plot") + scale_color_manual(values = c("Alive" = "darkgreen", "Dead" = "#cc0000")) + scale_fill_manual(values = c("Alive" = "#d9ead3", "Dead" = "#f4cccc")) + coord_flip()
```

This violin plot illustrates the distribution of ages for two groups:
those who are "Alive" and those who are "Dead". The width of each violin
represents the density of data points at a given age. In summary, living
individuals in this dataset tend to be younger, mostly between 25-50,
while deceased individuals were mostly aged between 50-75 when they
passed away.

<br>

```{r, echo=FALSE, error=FALSE, message=FALSE}
ggplot(data=filter(df, (covid_test_result == "Positive")), aes(x=mortality, y=number_of_disease, color = mortality, fill = mortality)) + geom_violin(na.rm=TRUE) + geom_boxplot(fill = "white", width=0.1, na.rm=TRUE) + ylab("number_of_disease") + ggtitle("Number of Diseases Distribution by Mortality: Violin Plot") + scale_color_manual(values = c("Alive" = "darkgreen", "Dead" = "#cc0000")) + scale_fill_manual(values = c("Alive" = "#d9ead3", "Dead" = "#f4cccc")) + coord_flip()
```

The violin plot showcases the distribution of the number of diseases
among individuals based on their mortality status. Living individuals
(in green) primarily have zero diseases, with a small distribution
around one disease. In contrast, deceased individuals (in red) display a
more spread-out distribution, primarily around one to three diseases,
with some having up to around eight diseases.

<br>

```{r, echo=FALSE, error=FALSE, message=FALSE}
sample_df = sample(df, 10000, replace=TRUE)
```

```{r, echo=FALSE, error=FALSE, message=FALSE, warning=FALSE}
ggplot(data=filter(sample_df, (covid_test_result == "Positive")), aes(x = age, y = number_of_disease, color=mortality)) + geom_point(size=1,position = position_jitter(width = 0.2, height = 0.2)) + scale_color_manual(values = c("Alive" = "#0d9962", "Dead" = "#922e41")) + xlab("Age") + ylab("Number of Prior Diseases") + ggtitle("Scatterplot of Number of Prior Diseases to Age - by Mortality (For a random sample of 10000)")
```

The scatterplot, based on 10,000 sampled data, depicts the relationship
between age and the number of prior diseases, differentiated by
mortality status. Alive individuals (in green) predominantly have fewer
diseases across all age groups, with a concentration at zero diseases.
Deceased individuals (in red) tend to have a wider range of diseases,
with a notable increase in disease count as age advances, especially
past 50.

<br>

```{r, echo=FALSE, error=FALSE, message=FALSE, warning=FALSE}
ggplot(data=filter(sample_df, (covid_test_result == "Positive")), aes(x = age, y = number_of_disease, color=mortality)) + geom_point(size=1,position = position_jitter(width = 0.2, height = 0.2)) + scale_color_manual(values = c("Alive" = "#0d9962", "Dead" = "#922e41")) + xlab("Age") + ylab("Number of Prior Diseases") + ggtitle("Scatterplot of Number of Prior Diseases to Age - by Mortality (For a random sample of 10000)") + facet_wrap(~ mortality) + geom_smooth(method="lm")
```

We separated the scatterplot into two groups to help analyze it better.
The scatterplot visually separates data on age and the number of prior
diseases based on mortality status. For alive individuals (in green),
there's a mild positive correlation between age and diseases, but it's
less pronounced. In contrast, deceased individuals (in red) exhibit a
clearer upward trend, suggesting a stronger correlation between age and
the number of diseases they had.

<br><br>

#### ***1st Statistical Test - Hypothesis Testing***

<br>

$$
\rm H_0: P_{Age \geq 60} - P_{Age \leq 18} = 0
\\
\rm H_A: P_{Age \geq 60} - P_{Age \leq 18} > 0
$$

In order to test the stated null hypothesis we will conduct a 2-sample
test for equality of proportions without continuity correction via a
prop.test() to check if the proportions of deaths among the different
age groups are the same.

We start by focusing on data from people who have actually tested
positive for COVID-19 and break it down further to look closely at
different age groups: one for seniors aged 60 and over, and another for
kids and teens under 18. By doing this step-by-step, we're making
special sets of data just for these particular age groups among those
who have caught the virus. This way, we can really zero in on what's
happening with the oldest and youngest folks who are dealing with
COVID-19, giving us a clearer picture within the larger group of
everyone who's tested positive.

```{r}
sub_df_pos = df %>% filter(covid_test_result == 'Positive')
sub_df_60 = sub_df_pos %>% filter(age >= 60)
sub_df_18 = sub_df_pos %>% filter(age <= 18)
```

```{r, echo=FALSE, error=FALSE, message=FALSE, results=FALSE}
length(sub_df_pos$age)
n.age_60 = length(sub_df_60$mortality)
n.age_18 = length(sub_df_18$mortality)

x.age_60 = favstats(~age, data=filter(sub_df_60,mortality == "Dead"))$n
x.age_18 = favstats(~age, data=filter(sub_df_18,mortality == "Dead"))$n

n.age_60
n.age_18
x.age_60
x.age_18
```

<br> ***Prop-Test***

```{r}
prop.test(c(x.age_60, x.age_18), c(n.age_60, n.age_18), alternative='two.sided', correct=FALSE)
```

Test Results \\ P-value: 0.0000 \\ 95% confidence interval: [0.3801,
0.3886]

We are able to compute $P$-value of $2.2 \times e^{-16}$, which is the
lowest possible $P$-value. This value being less than the level of
significance of 0.05 implies a rejection of the null hypothesis.

Thanks to our large dataset, the prop.test() also computes a compact and
accurate 95% confidence interval of [0.3801, 0.3886].

Therefore, based on the data, we can infer that the proportion of COVID
deaths among the 60+ age group is significantly higher than the under 18
age group. Furthermore, we can be 95% confident that the proportion of
60+ COVID deaths is greater than the sub 18 age group by 0.3801-0.3886.
This encompasses the observed difference in mortality rates of 0.3844.

```{r, echo=FALSE, error=FALSE, message=FALSE}
data.60 = c(rep(0, n.age_60-x.age_60), rep(1, x.age_60))
data.18 = c(rep(0, n.age_18-x.age_18), rep(1, x.age_18))

condition = c(rep(c(">= 60"), n.age_60), rep(c("<= 18"), n.age_18))
used = c(data.60, data.18)

mortality = data.frame(condition, used)

mortality60 = filter(mortality, condition==">= 60")
mortality18 = filter(mortality, condition=="<= 18")

phat.60 = numeric(1000)
phat.18 = numeric(1000)
phat.60_18 = numeric(1000)

for(i in 1:1000){
  phat.60[i] = sum(sample(mortality60$used, n.age_60, replace=TRUE))/n.age_60
  phat.18[i] = sum(sample(mortality18$used, n.age_18, replace=TRUE))/n.age_18
  
  phat.60_18[i] = phat.60[i] - phat.18[i]
}
phat.60_18.df = data.frame(phat.60, phat.18, phat.60_18)
```

Calculating the actual/observed difference in COVID mortality rates of
individuals 60 and over and individuals 18

```{r}
actual.60_18 = (x.age_60 / n.age_60) - (x.age_18 / n.age_18)
actual.60_18
```

<br>

To get a better picture of this confidence interval we have also decided
to compute a 95% confidence interval via bootstrapping.

The following shows a distribution for 1000 bootstraps difference in
mortality rate:

```{r, echo=FALSE, error=FALSE, message=FALSE}
ggplot(phat.60_18.df, aes(phat.60_18)) + geom_histogram(col="#45818e", fill= "#9ef2e4", ) + xlab("P{Age >= 60} - P{Age <= 18}") + ggtitle("Distribution for 1000 Bootstrap Difference in Mortality Rate \nP{Age >= 60} - P{Age <= 18}") + geom_vline(xintercept = actual.60_18, col="#e40909")
```

With this bootstrap method we can determine the 95% confidence intervals
for difference in proportions of mortality rates, mortality rates of the
60+ age group and sub 18 age group.

<br>

```{r, echo=FALSE, error=FALSE, message=FALSE, warning=FALSE}
ggplot(data=phat.60_18.df, aes(sample=phat.60_18)) + stat_qq(col="midnightblue") + stat_qqline(col="red") + ggtitle("Normal Probability Plot of Bootstrap value of Difference of mortality rate")
```

Based on a probability plot of differences computed, it appears to
follow a normal distribution as the data points in this plot are located
approximately along a straight line, it can be concluded that the
selected distribution correctly represents the data.

<br>

**95% confidence interval for Difference between**
$P_{Age \geq 60} - P_{Age \leq 18}$

```{r}
qdata(~phat.60_18, c(0.025,0.975), data=phat.60_18.df)
```

Based on this 95% confidence interval via bootstrapping we can be 95%
confident that the difference in COVID mortality rates of individuals 60
and over and individuals 18 and under is between 0.3802 and 0.3886. This
reaffirms the outcome obtained from the prop.test previously conducted
and the observed difference of 0.3844.

<br>

**95% confidence interval for** $P_{Age \geq 60}$

```{r}
qdata(~phat.60, c(0.025,0.975), data=phat.60_18.df)
```

Based on this 95% confidence interval via bootstrapping we can be 95%
confident that the COVID mortality rate of individuals 60 and over is
between 0.3986 and 0.4056. This confidence interval encompasses the
observed mortality rate of 31015/77113 = 0.4022.

<br>

**95% confidence interval for** $P_{Age \leq 18}$

```{r}
qdata(~phat.18, c(0.025,0.975), data=phat.60_18.df)
```

Based on this 95% confidence interval via bootstrapping we can be 95%
confident that the COVID mortality rate of individuals 18 and under is
between 0.0153 and 0.0203. This confidence interval encompasses the
observed mortality rate of 205/11520 = 0.0178.

Here we can visualize the differences in mortality rates between the
different age groups.

```{r, echo=FALSE, error=FALSE, message=FALSE, results=FALSE}
combined_df_60_18 = bind_rows(sub_df_60 %>% mutate(Dataset = "Age >= 60"),sub_df_18 %>% mutate(Dataset = "Age <= 18"))

head(combined_df_60_18,5)
```

```{r, echo=FALSE, error=FALSE, message=FALSE}
ggplot(data = combined_df_60_18, aes(x = Dataset, fill = mortality)) +
  geom_bar( position = "dodge", width = 0.6) +
  scale_fill_manual(values = c("Alive" = "#0d9962", "Dead" = "#922e41")) +
  ylab("Patient count") +
  xlab("Dataset") +
  ggtitle("Bar plot: Mortality vs Age groups") +
  theme_minimal()
```

```{r, echo=FALSE, error=FALSE, message=FALSE}
ggplot(data = combined_df_60_18, aes(x = Dataset, fill = mortality)) +
  geom_bar(position = "fill", width = 0.5) +
  scale_fill_manual(values = c("Alive" = "#0d9962", "Dead" = "#922e41")) + 
  ylab("Proportion") + 
  xlab("Dataset") +
  ggtitle("Stacked Bar Plot: Mortality vs Age Group (Proportions)") +
  theme_minimal()
```

<br><br>

#### ***2nd Statistisal Test - Chi-squared Test of Independence***

<br> $$
\rm H_0: \text{Co-Morbidities and the factors of interest are independent}
\\
\rm H_A: \text{Co-Morbidities and the factors of interest are dependent}
$$

As COVID has been heavily associated with co-morbidities, we want to
determine if there is any correlation between the number of diseases
among an individual and COVID attributes such as whether someone will
test positive for COVID and mortality. We will also explore the
correlation of co-morbidities with age and gender.

It is important to note that when conducting the following Chi-Square
tests for the whole dataset, the p-values for every test will fall well
below the 0.05 level of significance. This capability stems from our
extensive dataset, allowing us to calculate highly precise p-values and
confidence intervals, as exemplified in the preceding hypothesis test.
Consequently, R will approach a lower limit of 2.2e-15 for the computed
p-value. Likewise, enabling simulated p-values yields a consistent value
of 0.0004998 across all Chi-Square tests. Given this outcome, we've
chosen to investigate Chi-square test results using samples comprised of
500 data points.

<br>

**First let's take a look at whether a correlation exists between number
of diseases among an individual and status (testing positive for
COVID)**

$$
\rm H_0: \text{Number of Disease an individual has and their Covid Test Result are independent}
\\
\rm H_A: \text{Number of Disease an individual has and their Covid Test Result are not independent}
$$

<br> Contingency Table

```{r, echo=FALSE, error=FALSE, message=FALSE}

covid_test.df = subset(df, select = c(number_of_disease,covid_test_result))
covid_test.df = covid_test.df[complete.cases(covid_test.df), ]

sample.df = sample(covid_test.df, 500, replace=TRUE)
covid_test = tally(~number_of_disease + covid_test_result, data = sample.df)
covid_test
```

Chisq Result

```{r, echo=FALSE, error=FALSE, message=FALSE}
chi1 = chisq.test(covid_test, simulate.p.value=TRUE)
chi1

```

```{r, echo=FALSE, error=FALSE, message=FALSE, results=FALSE}
p.covid_test = chi1$p.value
test_stat.covid_test = chi1$statistic

p.covid_test
test_stat.covid_test
```

Given the $P$-value of 0.001499 being less than the significance level
of 0.05, we reject the null hypothesis. This means that based on the
COVID sample data, there is enough evidence to suggest a significant
association between the number of diseases an individual has and their
likelihood of testing positive for COVID.

<br>

We can visualize the COVID test results based on the number of diseases
as shown here.

```{r, echo=FALSE, error=FALSE, message=FALSE}
ggplot(data = covid_test.df, aes(x = number_of_disease, fill = covid_test_result)) +
  geom_bar( position = "dodge", width = 0.6) +
  scale_fill_manual(values = c("Positive" = "#efb914", "Negative" = "#432297")) +
  ylab("Patient count") +
  xlab("Number of Prior Diseases") +
  ggtitle("Bar plot: Covid Test Result vs Number of prior diseases") +
  theme_minimal()
```

```{r, echo=FALSE, error=FALSE, message=FALSE}
ggplot(data = covid_test.df, aes(x = number_of_disease, fill = covid_test_result)) +
  geom_bar(position = "fill", width = 0.6) +
  scale_fill_manual(values = c("Positive" = "#efb914", "Negative" = "#432297")) + 
  ylab("Proportion") + 
  xlab("Number of Prior Diseases") +
  ggtitle("Stacked Bar Plot: Covid Test Result vs Number of prior diseases (Proportions)") +
  theme_minimal()
```

<br><br>

**We take a look at whether a correlation exists between the number of
prior diseases among an individual and the severity of COVID symptoms**

$$
\rm H_0: \text{Number of Disease an individual has and Severity of COVID Symptoms are independent}
\\
\rm H_A: \text{Number of Disease an individual has and Severity of COVID Symptoms are not independent}
$$ <br> Contingency Table

```{r, echo=FALSE, error=FALSE, message=FALSE}
severity.df = subset(sub_df_pos, select = c(number_of_disease, covid_severity))
severity.df = severity.df[complete.cases(severity.df), ]

sample2.df = sample(severity.df, 500, replace=TRUE)
severity = tally(~number_of_disease + covid_severity, data = sample2.df)
severity = severity[, -4]
severity
```

Chisq Result

```{r, echo=FALSE, error=FALSE, message=FALSE}
chi2 = chisq.test(severity, simulate.p.value = TRUE)
chi2
```

```{r, echo=FALSE, error=FALSE, message=FALSE, results=FALSE}
p.severity = chi2$p.value
test_stat.severity = chi2$statistic

p.severity
test_stat.severity
```

Given the $P$-value of 0.3393 being more than the significance level of
0.05, we fail to reject the null hypothesis. This means that based on
the COVID sample data, there is not enough evidence to suggest a
significant association between the number of diseases an individual has
and their likelihood of testing positive for COVID.

In the context of this research, while the $P$-value for this particular
sample stands at more than 0.05, indicating a lack of statistical
significance, the $P$-value derived from the entire population data is
less than 0.05. Although the sample suggests no significant
relationship, the aggregate population data tells a different story,
highlighting a statistically significant association.

We can visualize the COVID Severity based on the number of diseases as
shown here. <br>

```{r, echo=FALSE, error=FALSE, message=FALSE}
ggplot(data = severity.df, aes(x = number_of_disease, fill = covid_severity)) +
  geom_bar( position = "dodge", width = 0.6) +

  ylab("Patient count") +
  xlab("Covid Severity") +
  ggtitle("Bar plot: Covid Severity vs Number of prior diseases") +
  theme_minimal()
```

```{r, echo=FALSE, error=FALSE, message=FALSE}
ggplot(data = severity.df, aes(x = number_of_disease, fill = covid_severity)) +
  geom_bar(position = "fill", width = 0.6) +
   
  ylab("Proportion") + 
  xlab("Covid Severity") +
  ggtitle("Stacked Bar Plot: Covid Severity vs Number of prior diseases (Proportions)") +
  theme_minimal()
```

<br><br>

**We examine whether there is a correlation between the number of
previous diseases in an individual and their Mortality Rate.**

$$
\rm H_0: \text{Number of Disease an individual has and Mortality are independent}
\\
\rm H_A: \text{Number of Disease an individual has and Mortality are not independent}
$$ <br> Contingency Table

```{r, echo=FALSE, error=FALSE, message=FALSE}
mortality.df = subset(sub_df_pos, select = c(number_of_disease, mortality))
mortality.df = mortality.df[complete.cases(mortality.df), ]

sample3.df = sample(mortality.df, 500, replace=TRUE)
mortality = tally(~number_of_disease + mortality, data = sample3.df)
mortality
```

Chisq Result

```{r, echo=FALSE, error=FALSE, message=FALSE}
chi3 = chisq.test(mortality, simulate.p.value = TRUE)
chi3

```

```{r, echo=FALSE, error=FALSE, message=FALSE, results=FALSE}
p.mortality = chi3$p.value
test_stat.mortality = chi3$statistic

p.mortality
test_stat.mortality
```

The computed $P$-value of 0.0004998 is also well below the 0.05
significance level threshold. This leads to the conclusion that we would
reject the null hypothesis. Thus, based on the COVID data, there is a
statistically significant association between the number of diseases an
individual has and their mortality. The number of diseases and mortality
are not independent; there exists a relationship between them.

We can visualize the Mortality based on the number of diseases as shown
here. <br>

```{r, echo=FALSE, error=FALSE, message=FALSE}
ggplot(data = mortality.df, aes(x = number_of_disease, fill = mortality)) +
  geom_bar( position = "dodge", width = 0.6) +
  scale_fill_manual(values = c("Alive" = "#0d9962", "Dead" = "#922e41")) +
  ylab("Patient count") +
  xlab("Mortality") +
  ggtitle("Bar plot: Mortality vs Number of prior diseases") +
  theme_minimal()
```

```{r, echo=FALSE, error=FALSE, message=FALSE}
ggplot(data = mortality.df, aes(x = number_of_disease, fill = mortality)) +
  geom_bar(position = "fill", width = 0.6) +
  scale_fill_manual(values = c("Alive" = "#0d9962", "Dead" = "#922e41")) + 
  ylab("Proportion") + 
  xlab("Mortality") +
  ggtitle("Stacked Bar Plot: Mortality vs Number of prior diseases (Proportions)") +
  theme_minimal()
```

<br><br>

**Now, we check whether there is a correlation between the number of
prior diseases among an individual and the Age range they belong to**

$$
\rm H_0: \text{Number of Disease an individual has and their Age Group are independent}
\\
\rm H_A: \text{Number of Disease an individual has and their Age Group are not independent}
$$ <br> Contingency Table

```{r, echo=FALSE, error=FALSE, message=FALSE}
age_range.df = subset(sub_df_pos, select = c(number_of_disease, age_range))
age_range.df = age_range.df[complete.cases(age_range.df), ]

sample4.df = sample(age_range.df, 500, replace=TRUE)
age_range = tally(~number_of_disease + age_range, data = sample4.df)
age_range
```

Chisq Result

```{r, echo=FALSE, error=FALSE, message=FALSE}
chi2 = chisq.test(age_range, simulate.p.value = TRUE)
chi2

```

```{r, echo=FALSE, error=FALSE, message=FALSE, results=FALSE}
p.age_range = chi2$p.value
test_stat.age_range = chi2$statistic

p.age_range
test_stat.age_range
```

Given the $P$-value is 0.0009995, which is well below the significance
level of 0.05, we would reject the null hypothesis. This means that
based on the COVID sample data, there is a statistically significant
association between the number of diseases an individual has and their
age group. In other words, age group and the number of diseases are not
independent; there is a relationship between them.

Here is a visualization that shows the counts and proportions of
different age ranges and the number of diseases an individual has among
those age groups. <br>

```{r, echo=FALSE, error=FALSE, message=FALSE}
ggplot(data = age_range.df, aes(x = number_of_disease, fill = age_range)) +
  geom_bar( position = "dodge", width = 0.6) +

  ylab("Patient count") +
  xlab("Number of Prior Diseases") +
  ggtitle("Bar plot: Age Range vs Number of prior diseases") +
  theme_minimal()
```

```{r, echo=FALSE, error=FALSE, message=FALSE}
ggplot(data = age_range.df, aes(x = number_of_disease, fill = age_range)) +
  geom_bar(position = "fill", width = 0.6) +
   
  ylab("Proportion") + 
  xlab("Number of Prior Diseases") +
  ggtitle("Stacked Bar Plot: Age Range vs Number of prior diseases (Proportions)") +
  theme_minimal()
```

<br><br>

**Finally, let's take a look at whether a correlation exists between
number of prior diseases among an individual and their Gender**

$$
\rm H_0: \text{Number of Disease an individual has and their Gender is independent}
\\
\rm H_A: \text{Number of Disease an individual has and their Gender is not independent}
$$ <br> Contingency Table

```{r, echo=FALSE, error=FALSE, message=FALSE}
sex.df = subset(sub_df_pos, select = c(number_of_disease, sex))
sex.df = sex.df[complete.cases(sex.df), ]

sample5.df = sample(sex.df, 500, replace=TRUE)
sex = tally(~number_of_disease + sex, data = sample5.df)
sex
```

Chisq Result

```{r, echo=FALSE, error=FALSE, message=FALSE}
chi4 = chisq.test(sex, simulate.p.value = TRUE)
chi4

```

```{r, echo=FALSE, error=FALSE, message=FALSE, results=FALSE}
p.sex = chi4$p.value
test_stat.sex = chi4$statistic

p.sex
test_stat.sex
```

Given the $P$-value of 0.7216 being more than the significance level of
0.05, we fail to reject the null hypothesis. This means that based on
the COVID sample data, there is not enough evidence to suggest a
significant association between the number of diseases an individual has
and their gender.

Here is a visualization that shows the counts and proportions of Gender
and the number of diseases an individual has. <br>

```{r, echo=FALSE, error=FALSE, message=FALSE}
ggplot(data = sex.df, aes(x = number_of_disease, fill = sex, group = sex)) +
  geom_bar( position = "dodge", width = 0.6) +
  scale_fill_manual(values = c("female" = "#cc3177", "male" = "#1263ac")) +
  ylab("Patient count") +
  xlab("Gender") +
  ggtitle("Bar plot: Gender vs Number of prior diseases") +
  theme_minimal()
```

```{r, echo=FALSE, error=FALSE, message=FALSE}
ggplot(data = sex.df, aes(x = number_of_disease, fill = sex, group = sex)) +
  geom_bar(position = "fill", width = 0.6) +
  scale_fill_manual(values = c("female" = "#cc3177", "male" = "#1263ac")) + 
  ylab("Proportion") + 
  xlab("Gender") +
  ggtitle("Stacked Bar Plot: Gender vs Number of prior diseases (Proportions)") +
  theme_minimal()
```

### **Conclusion**

The findings from both traditional proportion tests and bootstrap
methods indicate that individuals aged 60 and above have a significantly
higher mortality rate due to COVID-19 compared to individuals aged 18
and under. Both tests reaffirm the observed differences within the
dataset thanks to the tight 95% confidence intervals.

Our Chi-Square tests provided insights on comorbidities and other
factors of interest. With the entire dataset of around 1.5 million data
points at our disposal, we were able to compute a near zero $P$-value
across all Chi-Square tests. Given the substantial number of data points
we input into the chi-squared test, it is anticipated that it may detect
some form of association between the factors, even if this association
is influenced by external variables.

As anticipated, the results indeed revealed a significant link between
the quantity of pre-existing health conditions and several variables,
including COVID-19 test outcomes, COVID-19 severity, age, and gender.
Furthermore, a noteworthy connection was established between
comorbidities and the mortality rate, with mortality rates rising as the
number of comorbidities increased.

To generate more realistic insights, we reproduced our studies by
analyzing 500 random data points. While the majority of the results were
similar to those obtained from assessments of the complete dataset,
several variations led to different conclusions. Notably, our
investigation indicated that severity and gender appeared to be
uncorrelated with comorbidities. This finding, based on our sample
approach, lends credence to the claim that it constitutes a more
accurate conclusion. As we have seen, the large size of the dataset can
make Chi-square tests prone to consistently identify relationships. This
assumption is supported by the graphical representation, which clearly
indicates independence between gender and comorbidities, as well as
between comorbidities and severity. Consequently, our findings support
the notion that employing a sampling approach may enhance the accuracy
of Chi-square tests, particularly in contexts where the sensitivity of
these tests in large datasets can inadvertently indicate relationships.

Our study underscores the profound impact of age and comorbidities on
COVID-19 outcomes, emphasizing the need for targeted interventions,
especially for older individuals and those with multiple diseases.

\newpage

### **References**

-   Loza, A., Wong-Chew, R. M., Jiménez-Corona, M.-E., Zárate, S.,
    López, S., Ciria, R., Palomares, D., García-López, R., Iša, P.,
    Taboada, B., Rosales, M., Boukadida, C., Herrera-Estrella, A.,
    Mojica, N. S., Rivera-Gutierrez, X., Muñoz-Medina, J. E.,
    Salas-Lais, A. G., Sanchez-Flores, A., Vazquez-Perez, J. A., ...
    Gutiérrez-Ríos, R. M. (2023). Two-year follow-up of the COVID- 19
    pandemic in Mexico. Frontiers in Public Health, 10,
    1050673.https://doi.org/10.3389/fpubh.2022.1050673

-   Nizri, M. (2023). COVID-19 Dataset [Data set]. Kaggle. Retrieved
    October 17, 2023, from
    <https://www.kaggle.com/datasets/meirnizri/covid19-dataset>

-   Secretaría de Salud (Mexico). (2023). Información referente a casos
    COVID-19 en México [Data set]. Gobierno de México. Retrieved October
    17, 2023, from
    <https://datos.gob.mx/busca/dataset/informacion-referente-a-casos-covid-19-en-mexico>

\newpage

### **Appendix: All code for this report**

```{r ref.label=knitr::all_labels(), echo=TRUE, eval=FALSE}
```
